### 2013-11-09 article2013.Rmd
hieman epätoivoinen commit enne mikon tekstin lisäämistä

As Joseph Schumpeter (1942: 12) has once argued, "attitudes are coins that do not readily melt". This means that normative expectations, justice beliefs or welfare perceptions are very hard to change. In this way, welfare attitudes can be associated with policy changes by highlighting public sentiments, which are normative background assumptions or ideas that constrain decision-making and institutional change by limiting the range of programme that decision-making elites are likely to perceive as acceptable and legitimate to both their constituents and themselves (Campbell, 2004: 96-100). On the other hand, "the worlds on welfare" literature on welfare attitudes has in turn emphasised that type of welfare policy affects public attitudes (Andress and Heien, 2001; Arts and Gelissen, 2001; Jaeger, 2006; Svallfors, 1997). In regard to public perceptions of the causes of poverty, in the selective welfare-policy dominated liberal regimes explanations of poverty are more often based on individual morality that they are in the Continental and Nordic welfare states – where external and structural explanations prevail (Albrekt Larsen, 2006: 47-56; Alesina and Gleaser, 2004).

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. The development of post-socialist welfare state has received abundant scholarly attention over last years (e.g. Cerami and Vanhuysse, 2009, Inglot, 2008, Haggard and Kauffman, 2008; Cook, 2007; Aidukaite, 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasised the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 20 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka, 2009, Tomka, 2006, Cerami and Vanhuysse, 2009, Inglot, 2008).

Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasised clear differences between Western and Eastern European countries (Gallie and Paugam, 2002; Kallio and Niemelä, 2010; Lepianka, 2007; van Oorschot and Halman, 2000). East Europeans are more inclined to blame either the individual behaviour or the flaws of the economic system of country than their Western counterparts. However, there is a shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a detailed and more up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

Hence, utilising The Life in Transition Survey (LiTS) conducted jointly by the European Bank for Reconstruction and Development (EBRD) and the World Bank in 2010, the purpose of the article is to examine cross-country differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. The study applies multilevel techniques in order to analyse the role of individual and country-level explanatory factors as determinants of poverty attributions.

@@

Pioneering empirical studies on lay explanations of poverty emerged at the early 1970s when Joe R. Feagin (1972; 1975) examined Americans perceptions of the causes of poverty. He categorised reasons for poverty into three basic categories: individualistic, structural and fatalistic. Later, many factor analytic studies have given empirical support to Feagin’s categorisation (e.g. Feather, 1974; Furnham, 1982; Hunt, 1996; Morçöl, 1997; Nasser et al., 2002; Niemelä, 2008). Yet some of the recent studies have expanded our understanding by incorporating more contemporary beliefs into the attributional scales. These recent works emphasise especially the relevance of cultural attributions for poverty (Bullock et al., 2003; Cozzarelli et al., 2001).

van Oorschot and Halman (2000) have suggested a four-tier typology (Table 1) in which individual blame emphasises internal factors, such as undesirable behaviour of the poor, moral looseness and laziness. Individual fate type of explanation emphasises that poverty can be caused by uncontrollable and inescapable internal factors. These are factors such as personal misfortune, illness or just bad luck. Social fate points out that poverty is caused by societal factors, but these factors are due to impersonal and unavoidable processes such as economic recessions. Finally, social blame sees poverty as the result of processes induced and controlled by the actions of certain groups and parties in society. Therefore, these groups and parties can be blamed for poverty.

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the West and the East of Europe. Western Europeans are more inclined to support structural explanations for poverty than their Eastern counterparts. On the other hand, individual blame type of explanation is generally more popular in the East than in the West of Europe (Kallio and Niemelä, 2010; Kreidl, 2000; Lepianka, 2007; Lepianka et al., 2010; van Oorschot and Halman, 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the East Central European countries (Kallio and Niemelä, 2010). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the East of Europe than in Western European countries. Regarding the variation within the Eastern European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka, 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more the individuals themselves (Gorshkov and Tikhonova, 2006; Stephenson, 2000). These results are consistent with Lepianka's (2007: 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al., 1995).

Based on qualitative poverty studies in Central Asia, Caucasus and Ukraine and Moldova Dudwick et al. (2003: 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. They also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens.

@@

Besides the so called regime-effect, recent comparative studies on attributions for poverty have included country-level explanatory factors in the analysis (e.g. Albrekt Larsen, 2006; Gallie and Paugam, 2002; Lepianka, 2007; Lepianka et al., 2010). For instance, national-level economic conditions have been proved to be associated with welfare attitudes. Increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune, 2007). In regard to explanations of poverty, Gallie and Paugam (2002:21-24) found that in the majority of European countries, there was a marked decline of support for the individual laziness explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2010). They found that the larger the economic growth in a country is, the more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping. This study analyses the role of changes in economic performance during the period of global financial crises.

There are also considerable differences in socio-political context between the examined countries. The article analyses whether the level of income inequality and the state of democracy in a given country are associated with attributions for poverty. Regarding the state of democracy we can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues. The link between inequality and attributions for poverty is not that easy to predict. On the one hand, the public in countries with greater inequality is more inclined to mistrust "others" (e.g. Uslander, 2002). Therefore, the uneven distribution of wealth should mean greater support for the individualistic explanations for poverty. On the other hand, cross-national studies on attributions for poverty have shown that the support for the individualistic explanation is lower in the countries with lower level of income inequality (e.g. Gallie and Paugam, 2002; Kallio and Niemelä, 2010).

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalised by the individuals in the society in question (Taylor-Gooby, 1985; Svallfors, 1995). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice, 2001) and resource/class-based risks (Cusack et al. 2006; Blomberg et al. 2012; Finseraas and Ringdal, 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

Inspired by the studies which emphasise risks as the key mechanisms, this article tests to what extent factors related to financial hardship are associated – beyond the welfare regime – with perceptions of the causes of poverty. Following this line of reasoning we are interested in whether those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Thus, the research design in regard to determinants of poverty attributions enables us to study from a fairly unexplored perspective the different mechanisms through which individuals who have a greater risk to experience material destitution perceive the causes of poverty.

The previous studies on individual-level determinants of poverty perceptions have focused on individual’s socio-demographic characteristics (e.g. Bullock, 1999; Cozzarelli et al., 2001; Furnham, 1982; Morcöl, 1997; Niemelä, 2008). Results have, however, shown that socio-demographic variables explain very little and they do not reveal clear or consistent patterns. On the other hand, there are studies which support the assumption that those who perceive themselves to be poor are more likely to agree with the external cause of poverty than people well above the poverty line (Niemelä, 2008; Saunders, 2002: 155–156). Yet, there are also contradictory findings indicating that personal experience of poverty might lead to individualistic explanations. For instance, claimants of public welfare seem to hold negative views about other recipients (Bullock, 1999; 2004; Golding and Middleton, 1982: 178).

Studies which have explored the importance of individual’s economic situation on attributions for poverty have focused on individual’s cross-sectional situation of financial circumstances. However, as suggested in the literature on welfare attitudes not only the risk patterns but also the way people perceive their future risks might have an effect on attitudes toward the welfare state (Andress and Heien, 2001; Boeri et al., 2001; Blomberg et al., 2012; Nygaard Andersen and Ringdal, 2012). In their European comparison Blomberg et al. (2012) found that perceived future risks functioned as an intermediating factor between "objective" risk and welfare state attitudes, i.e. those who perceive their economic future as endangered were more in favour of state responsibility than those who do not expect their future situation to be difficult. Consequently, following the self-interest hypothesis, we can assume that those who perceive their future economic situation to be worse than now endorse more likely external reasons for poverty than those who do not expect difficulties in their future economic situation.

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin, 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyse past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyse whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to support external causes of poverty than those who has not experienced economic problems.

In addition to economic situation, other indicators which measure a greater than average risk of facing more permanent social problems, are "transfer dependency" and low education. Transfer dependency refers to a situation that the respondent’s main source of income relies on private or social transfers (Blomberg et al., 2012; Fridberg and Kangas, 2008). Because of their close relation with welfare system, it can be assumed that welfare recipients perceive more likely structural than individual reasons for poverty compared to the rest of the population.

Low education refers to the lack of capabilities of an individual. Prior findings do not reveal consistent patterns regarding the effects of education. On the one hand, some studies have found the so-called inverted U-relationship which means that people with a middle level of education support the individualistic explanation, whereas people with a low or a high level of education are more likely to endorse structural explanations (Feagin, 1972; cf. Furnham, 1982). Moreover, the relationship between education and fatalistic explanation has been found to be negative (Morcöl, 1997). On the other hand, there are studies which suggest that people with a lower level of education are more likely to favour individualistic explanations. Moreover, when the educational level increases, the likelihood of perceiving causes of poverty in individualistic terms decreases (Niemelä, 2008). It is therefore difficult to form any solid hypotheses as to whether low education as a social risk is associated with perceptions of the causes of poverty in a similar vein than other risk factors discussed above.

# Purpose of the study

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies have suggested (Kallio and Niemelä, 2010; Lepianka et al., 2010), there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space.

Hence, the general purpose of the study is to analyse public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries. Based on the findings of previous studies, we assume that in general post-socialist countries endorse more likely individual blame and social blame types of explanations than individual and social fate types of explanations. However, we hypothesise also that findings indicate relatively large variation between countries.

As a country-level determinants, we will analyse, whether this variation can be explained by different regimes within post-socialist countries and the post-socialist legacy. The study divides post-socialist countries into two groups: Central Eastern Europe (CEE) and CIS countries. The survey data enables us also to compare post-socialist countries to a selection of Western European nations. Thus, in order to examine the distinctiveness of post-socialist perception Western European countries are included in the descriptive analyses.  Besides the country groups we will also analyse the role of changes in country’s economic performance during the period of global financial crises, income inequality and the state of democracy.

As an individual-level determinants for the perceptions article test to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty, and whether these factors are associated with perceptions similar way regardless of the  groups of countries studied. Hereby, the research design aims at contributing to the theoretical reasoning regarding the importance of self-interest for attributions for poverty which assumes that those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks.

# Data and methods

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank (WB) in 2010. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample. More information regarding how households were sampled and people interviewed is available at ERBD, 2011. Second round of the survey had a special focus on how peoples lives were affected by the global economic crisis and it's aftermath. The dataset, documentation and publications are freely available at the survey website.

@@

@@

We are using LiTS2 data from 21 transition countries with additional 4 Western European countries (France, Germany, Italy and Sweden) (See Figure 1). In total 23513 households from 25 countries are included in the analysis. Transition countries were broken down into two groups. CEE includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania; and CIS[^****] Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia.

[^****]:Georgia withdrawn it's membership from CIS in 2008

## Dependent variables

The respondents were asked the following question which measures public perceptions of the causes of poverty: **In your opinion, what is the main reason why there are some people in need in our country today?** (Variable `q309` in LiTS-data)

Thus, the question is a standard forced-choice question which is used also in other comparative surveys such as in European Values Study and in Eurobarometer. According to van Oorschot and Halman (2000) different statements represent a four-tier typology of poverty explanations as follows: While the unlucky refers to individual fate, the laziness and lack of willpower represents individual blame. In addition, injustice in a society indicates social blame, whereas the view that poverty is simply part of modern life refers to social fate. The option 'other' was treated as missing data.

As our main interest here is in support for social and blame type of explanation and as we are employing logistic multivariate methods, the answers were recoded in two separate binomial variables. In \emph{social blame} variable we have  value 1 for supporting social blame type of explanation and 0 for supporting other type of explanation. Same logic applies to variable \emph{individual blame}.

## Individual level independent variables

Six individual-level risk predictors are included in the analysis. Variables are recoded as binomial where value 1 is coded for risk and 0 for absence of risk. The items included are described in the table 2.

@@

Table: Individual level predictors

-----------------------------------------------------------------------------
Variable                    Description
------------------------    -------------------------------------------------
Transfer dependency         Social or private transfers as main source of income

@@

Effect of financial crises  Whether respondents perceives that their household has suffered the financial crisis great or fair amount
-----------------------------------------------------------------------------

>>>>

As Joseph Schumpeter (1942: 12) once argued, "attitudes are coins that do not readily melt". This means that normative expectations, justice beliefs or welfare perceptions are very hard to change. In this way, welfare attitudes can be associated with policy changes by highlighting public sentiments, which are normative background assumptions or ideas that constrain decision-making and institutional change by limiting the range of programme that decision-making elites are likely to perceive as acceptable and legitimate to both their constituents and themselves (Campbell, 2004: 96-100). On the other hand, "the worlds on welfare" literature on welfare attitudes has in turn emphasized that type of welfare policy affects public attitudes (Andress and Heien, 2001; Arts and Gelissen, 2001; Jaeger, 2006; Svallfors, 1997). In regard to public perceptions of the causes of poverty, in the marginal welfare-policy dominated liberal regime, the poor will be asked to fulfil much harder deservingness criteria than in a universal social democratic regime (van Oorschot 2006). In countries, in which welfare programmes are more marginal and less generous, explanations of poverty are more often based on individual morality, reflecting the strength of the dominant individualistic ideology, than they are in the Continental and Nordic welfare states – where external and structural explanations prevail (Albrekt Larsen, 2006: 47-56; Alesina and Gleaser, 2004).

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. Since the early 1990s, the development of post-socialist welfare state has received abundant scholarly attention (e.g. Cerami and Vanhuysse, 2009, Inglot, 2008, Haggard and Kauffman, 2008; Cook, 2007; Aidukaite, 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasized the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 20 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka, 2009, Tomka, 2006, Cerami and Vanhuysse, 2009, Inglot, 2008).

The purpose of this article is to examine cross-country differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam, 2002; Kallio and Niemelä, 2013; Lepianka, 2007; van Oorschot and Halman, 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crises both at the individual and contextual level and their association with attributions for poverty. At the individual level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

The article is structured as follows: A brief overview of theoretical discourse on explanations of poverty and prior findings on cross-national differences in and contextual level determinants of attributions for poverty is presented in Section 2. Section 3 focuses on the self-interest approach as an individual level explanatory factor of the perceptions with particular interest of risk factors related to individual’s financial situation.  Research design, survey data and methods are described in detail in Section 4. The results of the study are presented in Section 5. Finally, the main conclusions, implications and limitations of the study are discussed in Section 6.

@@

Pioneering empirical studies on lay explanations of poverty emerged at the early 1970s when Joe R. Feagin (1972; 1975) examined Americans' perceptions of the causes of poverty. He categorized reasons for poverty into three basic categories: individualistic, structural and fatalistic. Later, many factor analytic studies have given empirical support to Feagin’s categorization (e.g. Feather, 1974; Furnham, 1982; Hunt, 1996; Morçöl, 1997; Nasser et al., 2002; Niemelä, 2008). Yet some studies have expanded our understanding by incorporating more contemporary beliefs into the attributional scales. These works emphasize especially the relevance of cultural attributions for poverty (Bullock et al., 2003; Cozzarelli et al., 2001).

van Oorschot and Halman (2000) have suggested a four-tier typology (Table 1) in which individual blame emphasizes internal factors, such as undesirable behaviour of the poor, moral looseness and laziness. Individual fate type of explanation emphasizes that poverty can be caused by uncontrollable and inescapable internal factors. These are factors such as personal misfortune, illness or just bad luck. Social fate points out that poverty is caused by societal factors, but these factors are due to impersonal and unavoidable processes such as economic recessions. Finally, social blame sees poverty as the result of processes induced and controlled by the actions of certain groups and parties in society. Therefore, these groups and parties can be blamed for poverty.

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the West and the East of Europe. Western Europeans are more inclined to support social explanations for poverty than their Eastern counterparts. On the other hand, individual blame type of explanation is generally more popular in the East than in the West of Europe (Kallio and Niemelä, 2013; Kreidl, 2000; Lepianka, 2007; Lepianka et al., 2010; van Oorschot and Halman, 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the East Central European countries (Kallio and Niemelä, 2013). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the East of Europe than in Western European countries. Regarding the variation within the Eastern European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka, 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

National case studies from Estonia and Russia have given support to findings of cross-national studies. There is strong support for the idea that the poor have only themselves to blame for their poverty. In addition, both of these countries have strong support for social blame as well. Estonians are as likely to blame the flaws of the economic system as the individualistic reasons. In Russia, on the other hand, the economic system is blamed even more than the individuals themselves (Gorshkov and Tikhonova, 2006; Stephenson, 2000). These results are consistent with Lepianka's (2007: 127–135) comparative analysis of Hungary, Russia, Czech Republic and Estonia. Interesting result is that social blame coincides strongly with individual blame. This illustrates the split-consciousness of public perceptions, which means the coexistence of both dominant and potentially challenging beliefs, where the former represents dominant stratification ideology or value structure of a given country and the latter everyday stratification-related experience of an individual (Kluegel et al., 1995).

Based on qualitative poverty studies in Central Asia, Caucasus and Ukraine and Moldova Dudwick et al. (2003: 23) states that, generally, people blamed their own poverty on the failure of the Soviet state and the corruption, indifference, and incompetence of their new leaders. At the same time, however, they often attributed the poverty of others to individual failure, such as laziness, alcoholism, having too many children to support, or having too few children to provide for their old age. Emphazising the positive association between horizontal and vertical trust (see e.g. Rothstein, 2000), they also reported that many of the respondents had become very distrustful of government officials and many suspected that government officials of exploiting the political and economic turmoil for their own gain, at the expense of ordinary citizens. With extensive country selection, the data utilized in this study provides a marvellous possibility to examine the association between attributions for poverty and the context regarding political process, civil liberties and political rights. We can assume that in those countries where the state of democracy is low, citizens blame more likely society for poverty than in those countries where citizens have political freedom and power to influence socio-political issues.

There are also considerable differences in other aspects of socio-political context between examined countries. One of these aspects is the extent of income inequality which varies extensively between examined countries. Thus, this study examines the role of income inequality as a contextual-level explanatory factor of public perceptions of the causes of poverty. The link between inequality and attributions for poverty is not that easy to predict. One the one hand, the public in countries with greater inequality is more inclined to mistrust “others” (e.g. Uslander, 2002). Therefore, the uneven distribution of wealth should mean stronger support for the individualistic explanation for poverty. On the other hand, cross-national studies on attributions for poverty have shown that the support for the individual blame is lower in countries with lower level of income inequality (e.g. Gallie and Paugam, 2002; Kallio and Niemelä, 2013).

@@

Besides the so called regime-effect, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crises in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crises, the research design of this article provides empirical evidence on the effects of crises to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune, 2007). In regard to explanations of poverty, Gallie and Paugam (2002:21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2013). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

@@

In regard to individual-level determinants, popular attitudes toward the welfare state have traditionally been assumed to be dependent on long-term class-related interest, short-term self-interest and the values and norms that have been internalized by the individuals in the society in question (Taylor-Gooby, 1985; Svallfors, 1995). More recent directions of comparative literature on welfare attitudes tend to also focus on the distribution of marketable capacities such as skill specificity (Iversen and Soskice, 2001) and resource/class-based risks (Cusack et al. 2006; Blomberg et al. 2012; Finseraas and Ringdal, 2012) as key to welfare attitudes. Thus, studies have assumed – and to some extent proved – that groups that have a greater risk of facing social problems or being dependent on welfare state benefits might perceive the government’s role in welfare issues in a distinct way compared to those who are not exposed to these risks.

Inspired by the studies which emphasise risks as the key mechanisms, this article tests to what extent factors related to financial hardship are associated – beyond the welfare regime – with perceptions of the causes of poverty. Following this line of reasoning, we are interested in whether those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Thus, the research design in regard to determinants of poverty attributions enables us to study from a fairly unexplored perspective the different mechanisms through which individuals who have a greater risk to experience material destitution perceive the causes of poverty.

The previous studies on individual-level determinants of poverty perceptions have focused on individual’s socio-demographic characteristics (e.g. Bullock, 1999; Cozzarelli et al., 2001; Furnham, 1982; Morcöl, 1997; Niemelä, 2008). Results have, however, shown that socio-demographic variables explain very little and they do not reveal clear or consistent patterns. On the other hand, there are studies which support the self-interest hypothesis that those who perceive themselves to be poor are more likely to blame society and social structures for poverty than people well above the poverty line (Niemelä, 2008; Saunders, 2002: 155–156). Yet, there are also contradictory findings indicating that personal experience of poverty might lead to individualistic explanations. For instance, claimants of public welfare seem to hold negative views about other recipients (Bullock, 1999; 2004; Golding and Middleton, 1982: 178).

Studies which have explored the importance of individual’s economic situation on attributions for poverty have focused on individual’s cross-sectional situation of financial circumstances. However, as suggested in the literature on welfare attitudes, not only the risk patterns but also the way people perceive their future risks might have an effect on attitudes toward the welfare state (Andress and Heien, 2001; Boeri et al., 2001; Blomberg et al., 2012; Nygaard Andersen and Ringdal, 2012). In their European comparison Blomberg et al. (2012) found that perceived future risks functioned as an intermediating factor between "objective" risk and welfare state attitudes, i.e. those who perceive their economic future as endangered were more in favour of state responsibility than those who do not expect their future situation to be difficult. Consequently, following the self-interest hypothesis, we can assume that those who perceive their future economic situation to be worse than now endorse more likely social blame type of explanation of poverty than those who do not expect difficulties in their future economic situation.

Not only the current situation or the perceived future but also the past experiences might be associated with individual’s evaluations (e.g. Kumlin, 2007). In regard to attributions for poverty, the effects of past experiences are – according to our knowledge – totally unexplored area of research. In this article we will analyse past experiences by two ways. First, we are interested in whether those whose current economic situation is lower than four years ago perceive the causes of poverty distinct way compared to those whose economic situation has improved or remained in the same level. Second, we will focus on the effects of the global financial crisis. We assume that economic crisis have affected especially those who are the most vulnerable groups in the society. Therefore, we will analyse whether those who report that the economic crisis has affected their household at least a fair amount perceive the causes of poverty differently than those who report that it has had no or only a little effect to their household. According to self-interest hypothesis, we assume that those whose economic situation has declined and those whose household has affected by economic crisis are more inclined to blame society for poverty than those who has not experienced economic problems.

In addition to economic situation, other indicators which measure a greater than average risk of facing more permanent social problems, are "transfer dependency" and low education. Transfer dependency refers to a situation that the respondent’s main source of income relies on private or social transfers (Blomberg et al., 2012; Fridberg and Kangas, 2008). Because of their close relation with welfare system, it can be assumed that welfare recipients perceive more likely social blame than indivual blame type of explanation of poverty compared to the rest of the population.

Low education refers to the lack of capabilities of an individual. Prior findings do not reveal consistent patterns regarding the effects of education. On the one hand, some studies have found the so-called inverted U-relationship which means that people with a middle level of education support the individualistic explanation, whereas people with a low or a high level of education are more likely to endorse structural explanations emphazising social inequalities and the flaws of the economic system of a given country (Feagin, 1972; cf. Furnham, 1982). On the other hand, there are studies which suggest that people with a lower level of education are more likely to favour individualistic explanations. Moreover, when the educational level increases, the likelihood of perceiving causes of poverty in individualistic terms decreases (Niemelä, 2008). It is therefore difficult to form any solid hypotheses as to whether low education as a social risk is associated with perceptions of the causes of poverty in a similar vein than other risk factors discussed above.

# Research design

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies have suggested (Kallio and Niemelä, 2013; Lepianka et al., 2010), there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries.

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank (WB) in 2010. The analyses cover 22 post-socialist countries and 4 Western European countries. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample (for more detailed description of sampling and data collection, see ERBD, 2011). The survey had a special focus on how people’s lives were affected by the global economic crises and its aftermath.

@@

The respondents were asked the following question which measures public perceptions of the causes of poverty: **In your opinion, what is the main reason why there are some people in need in our country today?**

Thus, the question is a standard forced-choice question which is used also in other comparative surveys such as in World Values Survey and in Eurobarometer. According to van Oorschot and Halman (2000) different statements represent a four-tier typology of poverty explanations as follows: While the unlucky refers to individual fate, the laziness and lack of willpower represents individual blame. In addition, injustice in a society indicates social blame, whereas the view that poverty is simply part of modern life refers to social fate (see Table 1). Based on the findings of previous studies, we assume that in general post-socialist countries endorse more likely individual blame and social blame types of explanations than individual and social fate types of explanations. However, we hypothesize also that findings indicate relatively large variation between countries.

As country-level determinants of perceptions, we will analyze, whether this variation can be explained by different regimes within post-socialist countries and the post-socialist legacy. The study divides post-socialist countries into two groups: Central Eastern Europe (CEE) and Commonwealth of Independent States (CIS). CEE includes Czech Republic, Estonia, Hungary, Bulgaria, Latvia, Lithuania, Poland, Slovakia, Slovenia, Romania; and CIS[^****] Armenia, Azerbaijan, Belarus, Georgia, Kazakhstan, Kyrgyzstan, Moldova, Tajikistan, Ukraine, Uzbekistan and Russia. The survey data enables us also to compare post-socialist countries to a selection of Western European nations (France, Germany, Italy and Sweden). Thus, in order to examine the distinctiveness of post-socialist perception, Western European countries are included in the descriptive analyses. (See Figure 1.)

@@

[^****]:Georgia withdrawn its membership from CIS in 2008

Besides the country groups we will also analyze the role of changes in country’s economic performance during the period of global financial crises, income inequality and the state of democracy. As described in previous sections, we assume that in those countries where the economic crises has meant declining economic conditions, public are more likely to endorse social explanations for poverty than in countries where national economy has continued to grow despite the global economic turbulence. Based on the theoretical discussion on civil liberties, political rights and institutional trust, we assume that the public support to social blame type of explanation is high in those countries where the state of democracy is low. Due to contradictory findings on the direction of the effect of income inequality, it is more difficult to provide any straightforward hypothesis on the importance of income inequality.

As an individual-level determinants for the perceptions article test to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty, and whether these factors are associated with perceptions similar way regardless of the  groups of countries studied. Hereby, the research design aims at contributing to the theoretical reasoning regarding the importance of self-interest for attributions for poverty which assumes that those who have financial difficulties or those who have a greater risk to be poor perceive the causes of poverty differently than those who are not exposed to these poverty-related risks. Analyses include six individual-level predictors, namely transfer dependency, low education, perceived low income level, income-level compared to the past and expected change in income in the future as well as perceived effect of the financial crises. Variables are recoded as binomial where value 1 is coded for risk and 0 for absence of risk. Both contextual and individual-level variables are described in the Table 2.

@@

Table: Description of independent variables in the analyses.

-----------------------------------------------------------------------------
Variable                    Description
------------------------    -------------------------------------------------
Individual-level

Transfer dependency         Social or private transfers as main source of income

@@

Effect of financial crises  Whether respondents perceives that their household has suffered the financial crisis great or fair amount

Country-level
